Algorithms for the Automated Correction of Vertical Drift in Eye-Tracking Data

### 2021-04-09 manuscript/manuscript.tex
Final changes to the manuscript before publication

\abstract{A common problem in eye tracking research is vertical drift---the progressive displacement of fixation registrations on the vertical axis that results from a gradual loss of eye tracker calibration over time. This is particularly problematic in experiments that involve the reading of multiline passages, where it is critical that fixations on one line are not erroneously recorded on an adjacent line. Correction is often performed manually by the researcher, but this process is tedious, time-consuming, and prone to error and inconsistency. Various methods have previously been proposed for the automated, post-hoc correction of vertical drift in reading data, but these methods vary greatly, not just in terms of the algorithmic principles on which they are based, but also in terms of their availability, documentation, implementation languages, and so forth. Furthermore, these methods have largely been developed in isolation with little attempt to systematically evaluate them, meaning that drift correction techniques are moving forward blindly. We document ten major algorithms, including two that are novel to this paper, and evaluate them using both simulated and natural eye tracking data. Our results suggest that a method based on dynamic time warping offers great promise, but we also find that some algorithms are better suited than others to particular types of drift phenomena and reading behavior, allowing us to offer evidence-based advice on algorithm selection.}

\keywords{algorithms; dynamic time warping; eye tracking; reading; vertical drift}

@@

Reading is a fundamental skill for navigating modern society and, as such, is subject to intense study in the cognitive and language sciences. Among the many tools that researchers use to investigate reading in the laboratory, eye tracking occupies a prominent position. Using this technique, participants' eye movements may be recorded as they read written material, providing a window into the relevant cognitive processes as they unfold. Technological advancements in eye tracking, particularly from the 1970s \parencite[see e.g.,][]{Rayner:1998}, have allowed researchers to collect increasingly accurate measures of eye movements during reading tasks, leading to great improvements in the investigation of the cognitive processes underlying reading and reading acquisition.

Many eye tracking studies involve the reading of single words or sentences. For example, researchers may embed target words into different sentence contexts and manipulate predictability \parencite[e.g.,][]{Rayner:2001}, display isolated words to gain insight into how a reader's eye moves when processing a word \parencite[e.g.,][]{Vitu:2004}, or reveal parts of words in a gaze-contingent fashion to investigate parafoveal processing \parencite[e.g.,][]{Schotter:2012}. Sentence reading experiments have also been essential in revealing the cognitive processes behind different levels of written language processing, from the width of the perceptual span \parencite[e.g.,][]{Rayner:1986,Blythe:2009} to the effects that word length and frequency have on eye movements \parencite[e.g.,][]{Joseph:2009,TiffinRichards:2015}, as well as the effects of syntactic \parencite[e.g.,][]{Frazier:1982,Pickering:1998} and lexical \parencite[e.g.,][]{Sereno:2006} ambiguity.

@@

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, and the degree and directionality of this error may fluctuate dynamically with each subsequent fixation, making it nontrivial to eliminate. Fig.~\ref{fig01}A depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig01_double_column.eps}}
	\vspace*{2pt}
	\caption{Example reading trials from an adult participant (left) and a child participant (right) taken from \textcite{Pescuma:0000}. Each dot represents a fixation and the size of the dots represents duration. The adult trial exhibits upward shift, especially in the lower left part of the passage. The child trial is extremely noisy and exhibits not just vertical drift issues but also many natural reading phenomena that will pose challenges to the algorithms.}
	\label{fig01}
	\end{figure}

Whatever the cause and however it manifests itself, vertical drift will ultimately have a negative impact on the analysis of eye tracking data because fixations will be mapped to words that were not actually being fixated at a given point in time (as we can see in Fig.~\ref{fig01}A). It is therefore incumbent on the researcher to recognize such issues when they occur and to take corrective measures. Specialized software packages, such as EyeLink Data Viewer (SR~Research, Toronto, Canada) or EyeDoctor (UMass Eyetracking Lab, Amherst, USA), provide the ability to manually move fixations, either individually or in small batches. However, manual realignment can be very time-consuming and is likely to be error-prone. In particular, the realignment process can be greatly complicated by other sources of noise or idiosyncratic reading behaviors. For example, Fig.~\ref{fig01}B depicts a reading trial by a child reader; in this case, not only is the data affected by drift issues, but there are also various natural reading behaviors, such as within- and between-line regressions, which add an additional layer of complexity to the task of realignment, not to mention the baseline level of noise and unusual features such as the arching sequence of fixations targeting line~4.

@@

To be clear, the algorithms we consider in this paper are restricted to one specific problem. Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014,Vadillo:2015}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see e.g.,][]{Palmer:2016,Nussli:2011}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

The paper proceeds in four main sections. First, we outline the algorithms. Second, we test the algorithms on simulated fixation sequences afflicted with various types of measurement error. Third, we test the algorithms on an eye tracking dataset (two examples from which are presented in Fig.~\ref{fig01}). And finally, we discuss the major properties of the algorithms, provide guidance to researchers about their use, and suggest ways in which they can be improved further. All code and data required to reproduce the analyses reported in this paper, as well as Matlab/Octave, Python, and R implementations of the algorithms, are available from the public data archive associated with this paper: https://doi.org/10.17605/OSF.IO/7SRKG

@@

The \texttt{attach} algorithm is the simplest of the algorithms considered in this paper. The algorithm simply attaches each fixation to its closest line. While this has the benefit of being extremely simple, it is generally not resilient to the kinds of drift phenomena described above. However, \texttt{attach} serves as a useful baseline algorithm, since it essentially corresponds to an eye tracking analysis in which no correction was performed---a standard analysis of eye tracking data would simply map fixations to the closest words or other areas of interest. We return to this point later in the paper.

@@

The \texttt{stretch} algorithm is loosely based on the the method proposed by \textcite{Lohmeier:2015} and shares some similarities with \textcite{MartinezGomez:2012} and \textcite{Nussli:2011}. \possessivecite{Lohmeier:2015} original method was designed for the reading of source code and therefore takes advantage of the fact that code has very irregular line lengths and indentation levels. The method works by finding an \textit{x}-offset, \textit{y}-offset, and scaling factor that, once applied to the fixations, minimizes alignment error between the fixations and lines of text.

@@

The final algorithm we consider, \texttt{warp}, is novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978}. DTW was used by the \texttt{compare} algorithm to provide a measure of dissimilarity between a gaze line and a text line. To our knowledge, however, there have been no previous reports of DTW being used directly to align fixations to text lines. This is somewhat surprising because DTW is the natural computational choice for tackling drift and alignment problems. The closest previously-described method is \textcite{Carl:2013}, who uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. This is quite complex, however, and we consider it to be a special case of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye tracking data.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig04_double_column.eps}}
	\vspace*{2pt}
	\caption{Mean accuracy of the ten algorithms in response to the five eye tracking phenomena. For example, some algorithms (\texttt{attach}, \texttt{chain}, \texttt{cluster}, \texttt{merge}, \texttt{regress}, and \texttt{stretch}) are adversely affected as the noise level is increased, while the other algorithms are either resilient to noise (\texttt{compare} and \texttt{split}) or entirely invariant to noise (\texttt{segment} and \texttt{warp}). The plotted lines have been vertically staggered to aid visualization.}
	\label{fig04}
	\end{figure}

@@

Results for the noise distortion parameter are shown in Fig.~\ref{fig04}A. Under zero noise, all algorithms perform at 100\% accuracy, but six of the algorithms are adversely affected by noise when it reaches a sufficiently high level of around 10: Of these, \texttt{chain} performs best, closely followed by \texttt{attach}, then \texttt{cluster}, \texttt{regress}, and \texttt{stretch}, and finally \texttt{merge}. Of the remaining algorithms, \texttt{compare} and \texttt{split} are highly resilient to noise, while \texttt{segment} and \texttt{warp} are entirely invariant.

@@

In terms of slope distortion (Fig.~\ref{fig04}B), when the parameter is set to zero, all algorithms perform perfectly, but as the slope becomes more extreme (in either the upward or the downward direction), five of the algorithms experience a sustained loss in accuracy. Of these, \texttt{cluster} and \texttt{stretch} generally perform best and, initially at least, \texttt{attach} performs worst; \texttt{chain} and \texttt{split} initially perform better than \texttt{attach}, but are eventually outperformed. Interestingly, although \texttt{regress} is mostly resilient to slope, it has two weak spots around the values of $-0.03$ and $0.03$. When the slope takes one of these values, \texttt{regress} struggles to disambiguate between (a) zero offset combined with the appropriate slope and (b) a large offset combined with slope in the opposite direction; if it selects the wrong option, fixations on one half of the passage will be misaligned, causing a substantial drop in accuracy. This reveals a hidden weakness of the \texttt{regress} algorithm, and we will see an example of it later. Of the remaining algorithms, \texttt{compare} and \texttt{merge} are highly resilient to slope, while \texttt{segment} and \texttt{warp} are invariant.

@@

In terms of shift (Fig.~\ref{fig04}C), when the parameter is zero, all algorithms perform perfectly, but as it becomes more extreme, five of the algorithms---\texttt{attach}, \texttt{chain}, \texttt{regress}, \texttt{split}, and \texttt{stretch}---drop in accuracy. In fact, \texttt{attach}, \texttt{chain}, and \texttt{split} produce identical results in the case of shift because they are all fundamentally reliant on absolute position. Somewhat surprisingly, \texttt{stretch} does not perform especially well on shift. This is because \texttt{stretch} can only handle up~to one full line of shift; any more than this and the bounds have to be relaxed, but this results in an objective function with multiple maxima which is difficult to optimize. The \texttt{compare} algorithm is mostly resilient to shift, while the remaining four algorithms---\texttt{cluster}, \texttt{merge}, \texttt{segment}, and \texttt{warp}---are invariant.

@@

Results for within-line regressions are shown in Fig.~\ref{fig04}D. When there are no within-line regressions, all algorithms perform at 100\%, but three of the algorithms drop off as the probability of within-line regression is increased. Of these, \texttt{compare} and \texttt{segment} track each other quite closely because they rely on identifying the return sweeps; \texttt{merge} is generally quite resilient, except when the parameter is around 0.7--0.9 because these values cause a large number of progressive sequences to be generated which cannot then be merged very freely, so the merge process tends to get trapped in local minima (i.e., bad mergers that happen early~on cannot later be reverted). Of the remaining algorithms, \texttt{split}\footnote{Unlike \texttt{compare} and \texttt{segment}, even if \texttt{split} misidentifies a regression as a return sweep, it will still be able to map the resulting gaze line to the appropriate text line because it assigns based on position rather than order.} and \texttt{warp} are highly resilient, while \texttt{attach}, \texttt{chain}, \texttt{cluster}, \texttt{regress}, and \texttt{stretch} are invariant.

@@

In terms of between-line regressions (Fig.~\ref{fig04}E), four algorithms are negatively impacted by increases in this parameter. Of these, \texttt{compare} and \texttt{split} can in principle find more than $m$ gaze lines, but they have difficulties identifying when a between-line regression occurs, while \texttt{segment} and \texttt{warp} are limited to identifying exactly $m$ gaze lines in strictly sequential order, so they fundamentally cannot handle between-line regressions. Of the remaining algorithms, \texttt{merge} is resilient to between-line regressions, while \texttt{attach}, \texttt{chain}, \texttt{cluster}, \texttt{regress}, and \texttt{stretch} are entirely invariant.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig05_single_column.eps}}
	\vspace*{2pt}
	\caption{Mean accuracy of the algorithms for each of the eye tracking phenomena. Darker cells indicate phenomena that an algorithm performs well on. A checkmark indicates that the algorithm is entirely invariant to the phenomenon in question, scoring 100\% in all 5000 simulations.}
	\label{fig05}
	\end{figure}

In this section, we have simulated five eye tracking phenomena that are particularly relevant to understanding the performance characteristics of the algorithms. Fig.~\ref{fig05} summarizes how accurately the algorithms perform on each phenomenon. No single algorithm is invariant---or even resilient---to all phenomena, although \texttt{merge} and \texttt{warp} come quite close: \texttt{merge} is only weak on noise, while \texttt{warp} is only weak on between-line regressions. In general, there tends to be a tradeoff between how well an algorithm can handle distortion and how well it can handle regressions; the ability to deal with one tends to come at the cost of the other. Nevertheless, in real world scenarios, performance will very much depend on the degree and relative prevalence of the phenomena. Furthermore, there are likely to be other important forms of measurement error and reading behavior that we have neglected to consider here, and those that we have considered are likely to interact in complex, unpredictable ways. It is therefore important to test the algorithms against natural eye tracking data to get a more holistic understanding of their performance.

@@

In this section, we test the algorithms against an eye tracking dataset that has been manually corrected by human experts. Unlike the simulations, there is no ground truth, and we cannot isolate particular phenomena; however, the benefit of this approach is that the phenomena are combined in a realistic way, allowing us to estimate how well the algorithms are likely to perform in real-world scenarios.

@@

We tested the algorithms on an eye tracking dataset collected by \textcite{Pescuma:0000}, which includes reading data for both adults and children, allowing us to test the algorithms on two distinct populations. Our general approach is illustrated in Fig.~\ref{fig06} and discussed over the following sections.

@@

\textcite{Pescuma:0000} collected eye tracking data for 12 passages from Italian children's stories (e.g., a passage from \textit{Goldilocks}). The passages were around 130 words in length, spanning 10--13 lines and were presented in 20-point Courier New (each character occupying around 0.45 degrees of visual angle). Either of two sets, each comprised of six passages, was administered for silent reading to a large sample of children aged 8--11 ($N=140$) and a smaller sample of adult controls ($N=33$) for a total of 877 reading trials. Eye movements were recorded using a tower mounted EyeLink 1000 Plus eye tracker (SR~Research, Toronto, Canada) for which a typical accuracy of 0.25--0.50 degrees is reported by the manufacturer. Recording was monocular (right eye) with a 1000~Hz sampling rate.

@@

Since is was impractical to manually correct all 877 trials (to do so would require months of work), we selected a sample for manual correction. For each of the 12 passages, we selected two reading trials by adult participants and two reading trials by child participants, for a total of 48 trials (5.5\% of the full dataset). The reading trials were selected pseudorandomly such that no single participant was represented more than once. Additionally, we manually checked and adjusted the sample to ensure it contained an equal balance of easy and challenging cases, as well as examples of all the various eye tracking phenomena discussed previously.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig10_double_column.eps}}
	\vspace*{2pt}
	\caption{Proportion of trials that surpassed (A) 90\%, (B) 95\%, and (C) 99\% accuracy, and (D) the proportion of corrections deemed acceptable by two human raters. The dark and light bars represent the adult and child datasets respectively.}
	\label{fig10}
	\end{figure}

Fig.~\ref{fig10} reports the proportion of corrections that surpassed an accuracy level of 90\%, 95\%, and 99\% by algorithm. If you are willing to accept relatively low accuracy at the trial level (e.g., 90\%, Fig.~\ref{fig10}A), then \texttt{cluster}, \texttt{merge}, and \texttt{stretch} will provide the best performance---a large proportion of corrections will meet this criterion. In comparison, if you have more stringent accuracy requirements at the trial level (e.g., 99\%, Fig.~\ref{fig10}C), then \texttt{segment}, \texttt{split}, and \texttt{warp} are likely to provide better performance. Of course, the cost of a more stringent accuracy criterion is that fewer corrections will be usable overall, and it is not possible to know which trials have low accuracy in the absence of manual correction data. This highlights the fact that it is currently not possible to confidently achieve a high level of accuracy in a high proportion of trials, so researchers may still need to invest a significant amount of time if a high level of accuracy is demanded.

To estimate how usable the corrections are likely to be to a typical researcher, we performed a more subjective analysis of their quality. All 480 algorithmic corrections were presented blind and in random order to two raters (JWC and VNP), who independently classified every correction as either ``acceptable'' or ``needs more work.'' The raters did not discuss in advance what criteria they would use to make these judgments, but agreement was nevertheless very high at 94\%. In addition to the overall number of errors, the raters weighed up other factors, such as how the errors were distributed over the passage, how challenging the input data seemed to be, and what effect the errors might have for downstream analyses. The results, which are shown in Fig.~\ref{fig10}D, suggest that \texttt{cluster}, \texttt{merge}, and \texttt{stretch} are likely to produce very satisfactory results on adult data.

>>>>

\abstract{A common problem in eye-tracking research is vertical drift---the progressive displacement of fixation registrations on the vertical axis that results from a gradual loss of eye tracker calibration over time. This is particularly problematic in experiments that involve the reading of multiline passages, where it is critical that fixations on one line are not erroneously recorded on an adjacent line. Correction is often performed manually by the researcher, but this process is tedious, time-consuming, and prone to error and inconsistency. Various methods have previously been proposed for the automated, post-hoc correction of vertical drift in reading data, but these methods vary greatly, not just in terms of the algorithmic principles on which they are based, but also in terms of their availability, documentation, implementation languages, and so forth. Furthermore, these methods have largely been developed in isolation with little attempt to systematically evaluate them, meaning that drift correction techniques are moving forward blindly. We document ten major algorithms, including two that are novel to this paper, and evaluate them using both simulated and natural eye-tracking data. Our results suggest that a method based on dynamic time warping offers great promise, but we also find that some algorithms are better suited than others to particular types of drift phenomena and reading behavior, allowing us to offer evidence-based advice on algorithm selection.}

\keywords{algorithms; dynamic time warping; eye tracking; line assignment; reading; vertical drift}

@@

Reading is a fundamental skill for navigating modern society and, as such, is subject to intense study in the cognitive and language sciences. Among the many tools that researchers use to investigate reading in the laboratory, eye tracking occupies a prominent position. Using this technique, participants' eye movements may be recorded as they read written material, providing a window into the relevant cognitive processes as they unfold. Technological advancements in eye tracking, particularly from the 1970s \parencite[see, e.g.,][]{Rayner:1998}, have allowed researchers to collect increasingly accurate measures of eye movements during reading tasks, leading to great improvements in the investigation of the cognitive processes underlying reading and reading acquisition.

Many eye-tracking studies involve the reading of single words or sentences. For example, researchers may embed target words into different sentence contexts and manipulate predictability \parencite[e.g.,][]{Rayner:2001}, display isolated words to gain insight into how a reader's eye moves when processing a word \parencite[e.g.,][]{Vitu:2004}, or reveal parts of words in a gaze-contingent fashion to investigate parafoveal processing \parencite[e.g.,][]{Schotter:2012}. Sentence reading experiments have also been essential in revealing the cognitive processes behind different levels of written language processing, from the width of the perceptual span \parencite[e.g.,][]{Rayner:1986,Blythe:2009} to the effects that word length and frequency have on eye movements \parencite[e.g.,][]{Joseph:2009,TiffinRichards:2015}, as well as the effects of syntactic \parencite[e.g.,][]{Frazier:1982,Pickering:1998} and lexical \parencite[e.g.,][]{Sereno:2006} ambiguity.

@@

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, and the degree and directionality of this error may fluctuate dynamically with each subsequent fixation, making it nontrivial to eliminate. Fig.~\ref{fig01}a depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig01_double_column.eps}}
	\vspace*{2pt}
	\caption{Example reading trials from (\textbf{a}) an adult participant and (\textbf{b}) a child participant taken from \textcite{Pescuma:0000}. Each dot represents a fixation and the size of the dots represents duration. The adult trial exhibits upward shift, especially in the lower left part of the passage. The child trial is extremely noisy and exhibits not just vertical drift issues but also many natural reading phenomena that will pose challenges to the algorithms.}
	\label{fig01}
	\end{figure}

Whatever the cause and however it manifests itself, vertical drift will ultimately have a negative impact on the analysis of eye-tracking data because fixations will be mapped to words that were not actually being fixated at a given point in time (as we can see in Fig.~\ref{fig01}a). It is therefore incumbent on the researcher to recognize such issues when they occur and to take corrective measures. Specialized software packages, such as EyeLink Data Viewer (SR~Research, Toronto, Canada) or EyeDoctor (UMass Eyetracking Lab, Amherst, USA), provide the ability to manually move fixations, either individually or in small batches. However, manual realignment can be very time-consuming and is likely to be error-prone. In particular, the realignment process can be greatly complicated by other sources of noise or idiosyncratic reading behaviors. For example, Fig.~\ref{fig01}b depicts a reading trial by a child reader; in this case, not only are the fixations affected by drift issues, but there are also various natural reading behaviors, such as within- and between-line regressions, which add an additional layer of complexity to the task of realignment, not to mention the baseline level of noise and unusual features such as the arching sequence of fixations targeting line~4.

@@

To be clear, the algorithms we consider in this paper are restricted to one specific problem. Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see, e.g.,][]{Zhang:2011,Zhang:2014,Vadillo:2015}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see, e.g.,][]{Palmer:2016,Nussli:2011}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see, e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

The paper proceeds in four main sections. First, we outline the algorithms. Second, we test the algorithms on simulated fixation sequences afflicted with various types of measurement error. Third, we test the algorithms on an eye-tracking dataset (two examples from which are presented in Fig.~\ref{fig01}). And finally, we discuss the major properties of the algorithms, provide guidance to researchers about their use, and suggest ways in which they can be improved further. All code and data required to reproduce the analyses reported in this paper, as well as Matlab/Octave, Python, and R implementations of the algorithms, are available from the public data archive associated with this paper: https://doi.org/10.17605/OSF.IO/7SRKG

@@

The \texttt{attach} algorithm is the simplest of the algorithms considered in this paper. The algorithm simply attaches each fixation to its closest line. While this has the benefit of being extremely simple, it is generally not resilient to the kinds of drift phenomena described above. However, \texttt{attach} serves as a useful baseline algorithm, since it essentially corresponds to an eye-tracking analysis in which no correction was performed---a standard analysis of eye-tracking data would simply map fixations to the closest words or other areas of interest. We return to this point later in the paper.

@@

The \texttt{stretch} algorithm is loosely based on the method proposed by \textcite{Lohmeier:2015} and shares some similarities with \textcite{MartinezGomez:2012} and \textcite{Nussli:2011}. \possessivecite{Lohmeier:2015} original method was designed for the reading of source code and therefore takes advantage of the fact that code has very irregular line lengths and indentation levels. The method works by finding an \textit{x}-offset, \textit{y}-offset, and scaling factor that, once applied to the fixations, minimizes alignment error between the fixations and lines of text.

@@

The final algorithm we consider, \texttt{warp}, is novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978}. DTW was used by the \texttt{compare} algorithm to provide a measure of dissimilarity between a gaze line and a text line. To our knowledge, however, there have been no previous reports of DTW being used directly to align fixations to text lines. This is somewhat surprising because DTW is the natural computational choice for tackling drift and alignment problems. The closest previously-described method is \textcite{Carl:2013}, who uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. This is quite complex, however, and we consider it to be a special case of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye-tracking data.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig04_double_column.eps}}
	\vspace*{2pt}
	\caption{Mean accuracy of the ten algorithms in response to the five eye-tracking phenomena. For example, some algorithms (\texttt{attach}, \texttt{chain}, \texttt{cluster}, \texttt{merge}, \texttt{regress}, and \texttt{stretch}) are adversely affected as the noise level is increased, while the other algorithms are either resilient to noise (\texttt{compare} and \texttt{split}) or entirely invariant to noise (\texttt{segment} and \texttt{warp}). The plotted lines have been vertically staggered to aid visualization.}
	\label{fig04}
	\end{figure}

@@

Results for the noise distortion parameter are shown in Fig.~\ref{fig04}a. Under zero noise, all algorithms perform at 100\% accuracy, but six of the algorithms are adversely affected by noise when it reaches a sufficiently high level of around 10: Of these, \texttt{chain} performs best, closely followed by \texttt{attach}, then \texttt{cluster}, \texttt{regress}, and \texttt{stretch}, and finally \texttt{merge}. Of the remaining algorithms, \texttt{compare} and \texttt{split} are highly resilient to noise, while \texttt{segment} and \texttt{warp} are entirely invariant.

@@

In terms of slope distortion (Fig.~\ref{fig04}b), when the parameter is set to zero, all algorithms perform perfectly, but as the slope becomes more extreme (in either the upward or the downward direction), five of the algorithms experience a sustained loss in accuracy. Of these, \texttt{cluster} and \texttt{stretch} generally perform best and, initially at least, \texttt{attach} performs worst; \texttt{chain} and \texttt{split} initially perform better than \texttt{attach}, but are eventually outperformed. Interestingly, although \texttt{regress} is mostly resilient to slope, it has two weak spots around the values of $-0.03$ and $0.03$. When the slope takes one of these values, \texttt{regress} struggles to disambiguate between (a) zero offset combined with the appropriate slope and (b) a large offset combined with slope in the opposite direction; if it selects the wrong option, fixations on one half of the passage will be misaligned, causing a substantial drop in accuracy. This reveals a hidden weakness of the \texttt{regress} algorithm, and we will see an example of it later. Of the remaining algorithms, \texttt{compare} and \texttt{merge} are highly resilient to slope, while \texttt{segment} and \texttt{warp} are invariant.

@@

In terms of shift (Fig.~\ref{fig04}c), when the parameter is zero, all algorithms perform perfectly, but as it becomes more extreme, five of the algorithms---\texttt{attach}, \texttt{chain}, \texttt{regress}, \texttt{split}, and \texttt{stretch}---drop in accuracy. In fact, \texttt{attach}, \texttt{chain}, and \texttt{split} produce identical results in the case of shift because they are all fundamentally reliant on absolute position. Somewhat surprisingly, \texttt{stretch} does not perform especially well on shift. This is because \texttt{stretch} can only handle up~to one full line of shift; any more than this and the bounds have to be relaxed, but this results in an objective function with multiple maxima which is difficult to optimize. The \texttt{compare} algorithm is mostly resilient to shift, while the remaining four algorithms---\texttt{cluster}, \texttt{merge}, \texttt{segment}, and \texttt{warp}---are invariant.

@@

Results for within-line regressions are shown in Fig.~\ref{fig04}d. When there are no within-line regressions, all algorithms perform at 100\%, but three of the algorithms drop off as the probability of within-line regression is increased. Of these, \texttt{compare} and \texttt{segment} track each other quite closely because they rely on identifying the return sweeps; \texttt{merge} is generally quite resilient, except when the parameter is around 0.7--0.9 because these values cause a large number of progressive sequences to be generated which cannot then be merged very freely, so the merge process tends to get trapped in local minima (i.e., bad mergers that happen early~on cannot later be reverted). Of the remaining algorithms, \texttt{split}\footnote{Unlike \texttt{compare} and \texttt{segment}, even if \texttt{split} misidentifies a regression as a return sweep, it will still be able to map the resulting gaze line to the appropriate text line because it assigns based on position rather than order.} and \texttt{warp} are highly resilient, while \texttt{attach}, \texttt{chain}, \texttt{cluster}, \texttt{regress}, and \texttt{stretch} are invariant.

@@

In terms of between-line regressions (Fig.~\ref{fig04}e), four algorithms are negatively impacted by increases in this parameter. Of these, \texttt{compare} and \texttt{split} can in principle find more than $m$ gaze lines, but they have difficulties identifying when a between-line regression occurs, while \texttt{segment} and \texttt{warp} are limited to identifying exactly $m$ gaze lines in strictly sequential order, so they fundamentally cannot handle between-line regressions. Of the remaining algorithms, \texttt{merge} is resilient to between-line regressions, while \texttt{attach}, \texttt{chain}, \texttt{cluster}, \texttt{regress}, and \texttt{stretch} are entirely invariant.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig05_single_column.eps}}
	\vspace*{2pt}
	\caption{Mean accuracy of the algorithms for each of the eye-tracking phenomena. Darker cells indicate phenomena that an algorithm performs well on. A checkmark indicates that the algorithm is entirely invariant to the phenomenon in question, scoring 100\% in all 5000 simulations.}
	\label{fig05}
	\end{figure}

In this section, we have simulated five eye-tracking phenomena that are particularly relevant to understanding the performance characteristics of the algorithms. Fig.~\ref{fig05} summarizes how accurately the algorithms perform on each phenomenon. No single algorithm is invariant---or even resilient---to all phenomena, although \texttt{merge} and \texttt{warp} come quite close: \texttt{merge} is only weak on noise, while \texttt{warp} is only weak on between-line regressions. In general, there tends to be a tradeoff between how well an algorithm can handle distortion and how well it can handle regressions; the ability to deal with one tends to come at the cost of the other. Nevertheless, in real world scenarios, performance will very much depend on the degree and relative prevalence of the phenomena. Furthermore, there are likely to be other important forms of measurement error and reading behavior that we have neglected to consider here, and those that we have considered are likely to interact in complex, unpredictable ways. It is therefore important to test the algorithms against natural eye-tracking data to get a more holistic understanding of their performance.

@@

In this section, we test the algorithms against an eye-tracking dataset that has been manually corrected by human experts. Unlike the simulations, there is no ground truth, and we cannot isolate particular phenomena; however, the benefit of this approach is that the phenomena are combined in a realistic way, allowing us to estimate how well the algorithms are likely to perform in real-world scenarios.

@@

We tested the algorithms on an eye-tracking dataset collected by \textcite{Pescuma:0000}, which includes reading data for both adults and children, allowing us to test the algorithms on two distinct populations. Our general approach is illustrated in Fig.~\ref{fig06} and discussed over the following sections.

@@

\textcite{Pescuma:0000} collected eye-tracking data for 12 passages from Italian children's stories (e.g., a passage from \textit{Goldilocks}). The passages were around 130 words in length, spanning 10--13 lines and were presented in 20-point Courier New (each character occupying around 0.45 degrees of visual angle). Either of two sets, each comprised of six passages, was administered for silent reading to a large sample of children aged 8--11 ($N=140$) and a smaller sample of adult controls ($N=33$) for a total of 877 reading trials. Eye movements were recorded using a tower mounted EyeLink 1000 Plus eye tracker (SR~Research, Toronto, Canada) for which a typical accuracy of 0.25--0.50 degrees is reported by the manufacturer. Recording was monocular (right eye) with a 1000~Hz sampling rate.

@@

Since it was impractical to manually correct all 877 trials (to do so would require months of work), we selected a sample for manual correction. For each of the 12 passages, we selected two reading trials by adult participants and two reading trials by child participants, for a total of 48 trials (5.5\% of the full dataset). The reading trials were selected pseudorandomly such that no single participant was represented more than once. Additionally, we manually checked and adjusted the sample to ensure it contained an equal balance of easy and challenging cases, as well as examples of all the various eye-tracking phenomena discussed previously.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig10_double_column.eps}}
	\vspace*{2pt}
	\caption{Proportion of trials that surpassed (\textbf{a}) 90\%, (\textbf{b}) 95\%, and (\textbf{c}) 99\% accuracy, and (\textbf{d}) the proportion of corrections deemed acceptable by two human raters. The dark and light bars represent the adult and child datasets respectively.}
	\label{fig10}
	\end{figure}

Fig.~\ref{fig10} reports the proportion of corrections that surpassed an accuracy level of 90\%, 95\%, and 99\% by algorithm. If you are willing to accept relatively low accuracy at the trial level (e.g., 90\%, Fig.~\ref{fig10}a), then \texttt{cluster}, \texttt{merge}, and \texttt{stretch} will provide the best performance---a large proportion of corrections will meet this criterion. In comparison, if you have more stringent accuracy requirements at the trial level (e.g., 99\%, Fig.~\ref{fig10}c), then \texttt{segment}, \texttt{split}, and \texttt{warp} are likely to provide better performance. Of course, the cost of a more stringent accuracy criterion is that fewer corrections will be usable overall, and it is not possible to know which trials have low accuracy in the absence of manual correction data. This highlights the fact that it is currently not possible to confidently achieve a high level of accuracy in a high proportion of trials, so researchers may still need to invest a significant amount of time if a high level of accuracy is demanded.

To estimate how usable the corrections are likely to be to a typical researcher, we performed a more subjective analysis of their quality. All 480 algorithmic corrections were presented blind and in random order to two raters (JWC and VNP), who independently classified every correction as either ``acceptable'' or ``needs more work.'' The raters did not discuss in advance what criteria they would use to make these judgments, but agreement was nevertheless very high at 94\%. In addition to the overall number of errors, the raters weighed up other factors, such as how the errors were distributed over the passage, how challenging the input data seemed to be, and what effect the errors might have for downstream analyses. The results, which are shown in Fig.~\ref{fig10}d, suggest that \texttt{cluster}, \texttt{merge}, and \texttt{stretch} are likely to produce very satisfactory results on adult data.
